---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

We will use dplyr package for this analysis

```{r message = FALSE}
library(dplyr)
```

First we load the data

```{r}
data <- read.csv("C:/Users/User/Desktop/Data Science/Reproducible Research/Project Week 2/activity.csv")
data$date <- as.Date(data$date , format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

We will ignore the missing values in the dataset.

```{r}
m <- complete.cases(data)
dat <- data[m,]
```

Calculation of the total number of steps taken per day:
```{r}
da <- group_by(dat, date)
Q1 <- as.data.frame(summarize(da, totsteps = sum(steps)))
```

Histogram of the total number of steps taken each day
```{r}
hist(Q1$totsteps, xlab ="total steps", main = "Plot 1")
```

Mean and median of the total number of steps taken per day:
```{r message = FALSE, }
Q1.2 <- as.data.frame(summarize(da, totalsteps = sum(steps), mean = mean(steps), median = median(steps)))
Q1.2
```

## What is the average daily activity pattern?

We will ignore the missing values in the dataset.

```{r}
m <- complete.cases(data)
dat <- data[m,]
```

Make a plot of the 5 minute interval vs the average number of steps taken, averaged across all days.

```{r message = FALSE}
da2 <- group_by(dat, interval)
Q2 <- as.data.frame(summarize(da2, totsteps = sum(steps), mean = mean(steps), interval = interval))
with(Q2, plot(interval, mean, type = "l", main = "Plot 2", ylab = "average number of steps taken"))
```

The 5 minute interval on average across all days that contains the maximum number of steps is calculated with:

```{r}
print(Q2[which.max(Q2$mean), "interval"])
```


## Imputing missing values

The total number of missing values in the dataset is calculated with

```{r}
sum(is.na(data))
```

We create a new dataset called **copy_data** that is a copy of the initial data set.

```{r}
copy_data <- data.frame(data)
```

We fill in all the missing values of the dataset **copy_data**, with the mean for the 5 minute interval.

```{r message = FALSE}
pls <- group_by(copy_data, interval)
Q3 <- as.data.frame(summarize(pls, mean = mean(steps, na.rm = TRUE)))
copy_data$steps <- ifelse(is.na(copy_data$steps) == TRUE, Q3$mean[Q3$interval %in% copy_data$interval], copy_data$steps) 
```

We rearrange the data to obtain the total number of steps taken each day, and use it to make a histogram:

```{r message = FALSE}
gr <- group_by(copy_data, date)
sum3 <- as.data.frame(summarize(gr, totsteps = sum(steps)))
hist(sum3$totsteps, xlab = "total number of steps", main = "Plot 3")
```

We calculate the mean and median of the total number of steps taken per day:

```{r message = FALSE, }
Q3.2 <- as.data.frame(summarize(gr, totalsteps = sum(steps),  mean = mean(steps), median = median(steps)))
Q3.2
```


We can see that the values do not almost differ from the first part of the assigment where we removed the missing values.

While mean.x, median.x and totalsteps.x correspond to the dataset **Q3.2** whereas, mean.y, median.y and totalsteps.y correspond to the dataset **Q1.2** where the missing values where removed.

```{r}
last <- merge(Q3.2, Q1.2, by = "date", all = TRUE)
last
```


## Are there differences in activity patterns between weekdays and weekends?

We will use lattice package for this analysis

```{r message = FALSE}
library(lattice)
```

We create a new factor variable in the dataset with two leveles; weekday and weekend.

```{r}
q4 <- mutate(copy_data, day = weekdays(copy_data$date))
q4$day <- ifelse(q4$day %in% c("sábado", "domingo"), "weekend", "weekday")
head(q4)
```

We rearrange the dataset **copy_data**, to make a plot of the 5 minute interval vs the average number of steps taken, averaged across all weekday days or weekend days.

```{r message = FALSE}
gr4 <- group_by(q4, day, interval)
Q4.2 <- as.data.frame(summarize(gr4, mean = mean(steps)))
xyplot(mean ~ interval | day, data = Q4.2 , layout = c(1, 2), type = "l", main = "Plot 4", ylab = "average number of steps taken")
```
